Lock-free Skiplist-based Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-02-20 paper/paper.tex
Notes, concepts, definitions

In the past decade, advancements in computer performance have been made mostly
through increasing the number of processors instead of higher clock speeds.
This development necessitates new approaches to data structures and algorithms
that take advantage of concurrent execution on multiple threads and processors.

This paper focuses on the priority queue data structure, consisting of two operations
traditionally called \lstinline|Insert| and \lstinline|DeleteMin|. \lstinline|Insert|
places an item together with its priority into the queue, while \lstinline|DeleteMin|
removes and returns the highest priority item. Both operations are expected to have
a complexity of at most $O(\log n)$. Priority queues are used in a large variety
of situations such as shortest path algorithms and scheduling.

@@

In the following, we examine the evolution of concurrent priority queues from
an early prominent Heap-based algorithm through initial SkipList-based designs
to current state-

>>>>

% Move to many cores, why do we need concurrent algorithms/data structures?

In the past decade, advancements in computer performance have been made mostly
through an increasing number of processors instead of higher clock speeds.
This development necessitates new approaches to data structures and algorithms
that take advantage of concurrent execution on multiple threads and processors.

% The priority queue - semantics and usage summary.

This paper focuses on the priority queue data structure, consisting of two operations
traditionally called \lstinline|Insert| and \lstinline|DeleteMin|. \lstinline|Insert|
places an item into the queue together with its priority, while \lstinline|DeleteMin|
removes and returns the highest priority item. Both of these operations are expected to have
a complexity of at most $O(\log n)$. Priority queues are used in a large variety
of situations such as shortest path algorithms and scheduling. % See liu2012lock for more.

% Concurrent priority queues - heaps, trees, skiplists.

@@

% Our plan for the paper - somehow this kind of duplicates the last paragraph though.

In the following, we examine the evolution of concurrent priority queues. We begin
with first steps which used fine-grained locking to avoid the bottleneck of a single
global lock. Initial SkipList-based structures offered better disjoint-access
parallelism, and finally current state of the art lock-free priority queues minimize
contention in \lstinline|DeleteMin|.

% TODO: In section x, we ... Again, this partially duplicates the last paragraph.

\section{Concepts and Definitions}

% Blocking, lock-free and wait-free data structures.

Concurrent data structures are intended to be accessed simultaneously by several processes
at once. \emph{Lock-based} structures ensure that only a limited number of processes may enter
a critical section at once. \emph{Lock-free} data structures eschew the use of locks, and guarantee
that at least a single process makes progress at all times. Since lock-free structures are
non-blocking, they are not susceptible to priority inversion, deadlock, and livelock.
\emph{Wait-freedom} further guarantees that every process finishes each operation in a bounded number of steps. 
In practise, wait-freedom often introduces an unacceptable overhead; lock-freedom
however has proven to be both efficient and to scale well to large numbers of processes.

% Linearizability, sequential and quiescent consistency.

There are several different criteria which allows reasoning about the correctness of concurrent
data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
effect at a single instant in time at so-called linearization points. Linearizability is composable ---
any data structure composed of linearizable objects is also linearizable.
\emph{Sequential consistency} \cite{lamport1979make} requires the result of a set of operations
executed in parallel to be equivalent to the result of some sequential ordering of the same
operations.
\emph{Quiescently consistent} \cite{shavit1996diffracting}
guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
after a period of quiescence, i.e. an interval without active operations has passed.

% Disjoint-access parallelism.

\section{Fine-grained Locking Heaps}
\section{SkipList-based Priority Queues}
